formula1_start_vs_finish.Rmd

### 2024-03-19 inst/formula1_start_vs_finish.Rmd
improvements

```{r definemakeH3}
`makeH3` <- function(i,lambda){
  suppressWarnings(a <- setNames(as.numeric(finish[,1]),rownames(finish)))
  suppressWarnings(b <- setNames(as.numeric(grid  [,1]),rownames(finish)))
  wanted <- !is.na(b)
  a <- a[wanted]
  b <- b[wanted]
  a[is.na(a)] <- 0 # did not finish
  H <- ordervec2supp(a)
  for(i in seq_along(b)){H[names(b)[i]] <- as.weight(lambda^(i-1))}
  return(H)
}

likelam <- function(lambda){
  H <- hyper3()
  for(i in seq_len(ncol(finish))){H <- H + makeH3(1,lambda=lambda)}
  return(H)
}
```

```{r evaluatelamrange,cache=TRUE}

f <- function(lambda){
  H <- likelam(lambda)
  loglik(maxp(H,n=4),H)
}
f(0.9)
f(0.901)
```

```{r evalam,cache=TRUE}
lam <- seq(from=0.9,to=1.05,len=5)
like <- sapply(lam,f)
```

```{r plotlikelam}
plot(lam,like,type='b')
```

>>>>

```{r definereplacenonzerowithorder}
replacenonzerowithorder <- function(x){
  nonzero <- x!=0
  x[nonzero] <- rank(x[nonzero])
  return(x)
}

order_non_NA_NZ <- function(x){ # leaves NA and 0 entries unchanged; replaces other elements with their order
  wanted <- !is.na(x)
  x[wanted] <- replacenonzerowithorder(x[wanted])
  return(x)
}

topn <- 15
finish <- finish[seq_len(topn),]
options(use_alabama=TRUE)

`makeH3` <- function(i,lambda,topn,finish,grid){
  suppressWarnings(a <- setNames(as.numeric(finish[,i]),rownames(finish)))
  suppressWarnings(b <- setNames(as.numeric(grid  [,i]),rownames(finish)))
  wanted <- !is.na(b)
  a <- a[wanted]
  b <- b[wanted]
  a[is.na(a)] <- 0 # did not finish
  a <- a[seq_len(topn)]
  a <- order_non_NA_NZ(a)
  H <- ordervec2supp(a)
  for(j in seq_along(b)){
    if(!is.na(names(b)[j])){H[names(b)[j]] <- as.weight(lambda^b[j])}
  }
  return(H)
}

likelam <- function(lambda,topn,finish,grid){
  H <- hyper3()
  for(i in seq_len(ncol(finish))){H <- H + makeH3(i,lambda=lambda,topn=topn,finish=finish,grid=grid)}
  return(H)
}
```

```{r evaluatelambbda,cache=TRUE}
f <- function(lambda,topn,finish,grid){
  H <- likelam(lambda,topn,finish,grid)
  m <- maxp(H,n=1)
  return(c(loglik(m,H),lambda,m))
}
```

```{r evalamlam,cache=TRUE}
system.time(lam <- seq(from=0.85, to=1.01, len=15))
system.time(bigmat <- sapply(lam,f,topn=topn,finish=finish,grid=grid))
```

```{r showbigmat}
head(bigmat)
```

```{r max_alpha,cache=TRUE}
system.time(alpha_MLE <- optimize(function(a){f(a,topn=topn,finish=finish,grid=grid)[1]},c(0.95,0.99),maximum=TRUE))
alpha_MLE
```

```{r plotlikelambda}
like <- bigmat[1,]
like <- like - max(like)
plot(lam,like,type='b')
abline(h=c(0,-2))
abline(v=1)
segments(x0=0.9654,y0=-5,y1=0)
grid()
```

```{r showmaxalpha}
if(FALSE){print(alpha_MLE)}
```

```{r conventionalhyper2max,cache=TRUE}
H2 <- ordertable2supp(finish)
mH2 <- maxp(H2)
```

```{r}
dotchart(mH2,pch=16)
dotchart(log(mH2),pch=16)
```

# Above was 2023, now do the 2022 season

```{r loadfinish22}
setwd("~/rstudio/hyper2/inst")
finish <- read.table("formula1_2022.txt",header=TRUE)
finish22 <- finish[,seq_len(ncol(finish)-1)]
finish22[,1:5]
grid22 <- read.table("formula1_starting_2022.txt",header=TRUE)
grid22[,1:5]
```

```{r mlecalc22, cache=TRUE}
system.time(
    alpha_MLE22 <- optimize(function(a){f(a,topn=topn,finish=finish22,grid=grid22)[1]},c(0.95,0.99),maximum=TRUE)
    )
alpha_MLE22
```

```{r evalam22,cache=TRUE}
lam <- seq(from=0.90, to=1.05, len=15)
system.time(bigmat <- sapply(lam,f,topn=topn,finish=finish22,grid=grid22))
```

```{r showbigmat22}
head(bigmat)
```
